Exploring probability distributions for bivariate temporal granularities

### 2019-12-16 main.Rmd
data set changed from Ai, Bj to C1, C2

A recent tidy data structure to support exploration and modeling of temporal data is tsibble [@wang2019tsibble], where data is structured in a semantic manner with reference to observations and variables, with the time index stated explicitly. Since all cyclic granularities can be expressed in terms of the index set, we consider the data structure in (Figure \@ref(fig:data-structure) for exploration of temporal data of this kind. This is extending the columns of tsibble by including the cyclic granularities.Now suppose we want to explore the measurement variable $v$ across two cyclic granularities $C_1$ and $C_2$, such that $C_1$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_n$\}, and $C_2$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_m$\}. Also, let $S_{ij}$ be the set of index set such that for all $s \in S_{ij}$, $C_1(s) = A_i$ and $C_2(s) = B_j$. Data subsets like <$A_i$, $B_j$, $v(s)$> can be obtained for all $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$  which will lead to $nm$ such sets. These sets can form the basis for exploration and analysis of the measured variable across these bivariate cyclic granularities.

@@

The way cyclic granularities relate become important when we consider the data structure <$A_i$, $B_j(s)$, $v(s)$> $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$ for exploration. We will discuss few cases, where one or more of these $nm$ sets will be empty, which will essentially mean that the measured variable can not be explored for all combinations of the categories/levels of the cyclic granularities. To consider a general framework, we will define pairs of cyclic granularities as harmonies or clashes.

@@

Drawing from the grammar of graphics, if <$A_i$, $B_j(s)$, $v(s)$> $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$ serves as the basis of visualizing the distribution of the measured variable, the following layers can be specified:

- Data: <$A_i$, $B_j$, $v(s)$> $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$ 
- Aesthetic mapping: $A_i$ mapped to $x$ position and $v(s)$ to $y$ position
- Statistical transformation: Any descriptive or smoothing statistics that summarizes distribution of $v(s)$
- Geometric objects: Any geometry displaying distribution, for example, boxplot, letter value, violin, ridge or highest density region plots
- Facet:  $B_j$

<!-- To add additional variables to a plot, we can use other aesthetics like colour, shape, and size or facetting. Facetting creates tables of graphics by displaying the same graph for each subset. -->

@@

<!-- Aesthetics refers to the scales on to which we want to map our data. we create plots by layering grammarical elements on top of one another and use aesthetic maps -->

@@

<!-- To add additional variables to a plot, we can use other aesthetics like colour, shape, and size or facetting. Facetting creates tables of graphics by displaying the same graph for each subset. -->
<!-- - Facet:  $B_j(s)$ -->

@@

Choice of plots are dictated by the statistical transformations and geometric objects used for the visualization. The basic plot choice for our data structure is the one that can display densities. In Section \ref{sec:measured-variable}, it is dicussed that these densities could be estimated using Kernel density estimates or some descriptive statistics. The latter do not allow us to see the shape, skewness, nature of the tail or multi-modality with so much clarity as the former. However, they avoid much clutter and help us to focus on some specific properties of the data. Kernel density estimates provide more detailed information about the distribution but depend on smoothing parameters and hence can vary for the same data set. In practice, the densities are estimated reasonably with at least 30 observations. However, even with large sample size, choosing too large a value for bandwidth can cause the estimate to oversmooth the data. The summaries based on descriptive statistics remain free of any tuning parameters or Kernels, which is sometimes more desirable for exploration.

@@

When we consider the layers of grammars considered for our data structure <$A_i$, $B_j$, $v(s)$> $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$, mapping $A_i$ mapped to $x$ position and $B_j$ to facets
suggests that levels of $A_i$'s which are placed in close proximity are to be compared against each other and each level of $B_j$ represent a group. When $A_i$ mapped to facets and $B_j$ is mapped to $x$ axis, distributions are compared across levels of $B_j$ and levels if $A_i$ start acting as a group. However, the variables used in the layers are the same, the difference in mapping emphasizes different behavior of the variables.

>>>>

A recent tidy data structure to support exploration and modeling of temporal data is tsibble [@wang2019tsibble], where data is structured in a semantic manner with reference to observations and variables, with the time index stated explicitly. Since all cyclic granularities can be expressed in terms of the index set, we consider the data structure in (Figure \@ref(fig:data-structure) for exploration of temporal data of this kind. This is a two-dimensional array extending the columns of tsibble by including the cyclic granularities. Let us consider two cyclic granularities $C_1$ and $C_2$, such that $C_1$ maps index set to a set $\{A_1, A_2, A_3, \dots, A_n$\}, and $C_2$ maps index set to a set $\{B_1, B_2, B_3, \dots, B_m$\} and $v$ denotes the measurement variable. Data sets of the form <$C_1$, $C_2$, $v$> can then form the basis for exploration and analysis of the measured variable across bivariate cyclic granularities.

<!--  These subsets can form the basis for exploration and analysis of the measured variable across these bivariate cyclic granularities. -->

@@

The way cyclic granularities relate become important when we consider the data structure in \@autoref{fig:data-structure}. While considering data sets of the form $<C_1, C_2, v>$, let $S_{ij}$ be the set of index set such that for all $s \in S_{ij}$, $C_1(s) = A_i$ and $C_2(s) = B_j$. Data subsets like <$A_i$, $B_j$, $v(s)$> can be obtained for all $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$  which will lead to $nm$ data subsets. We will discuss few cases, where one or more of these $nm$ sets will be empty, which will essentially mean that the measured variable can not be explored for all combinations of the categories/levels of the cyclic granularities. To consider a general framework, we will define pairs of cyclic granularities as harmonies or clashes.

@@

Drawing from the grammar of graphics, if <$C_1$, $C_2$, $v$> serves as the basis of visualizing the distribution of the measured variable, the following layers can be specified:

- Data: <$C_1$, $C_2$, $v$>
- Aesthetic mapping: $C_1$ mapped to $x$ position and $v$ to $y$ position
- Statistical transformation: Any descriptive or smoothing statistics that summarizes distribution of $v(s)$
- Geometric objects: Any geometry displaying distribution, for example, boxplot, letter value, violin, ridge or highest density region plots
- Facet:  $C_2$

<!-- To add additional variables to a plot, we can use other aesthetics like colour, shape, and size or faceting. Faceting creates tables of graphics by displaying the same graph for each subset. -->

@@

<!-- Aesthetics refers to the scales on to which we want to map our data. we create plots by layering grammatical elements on top of one another and use aesthetic maps -->

@@

<!-- To add additional variables to a plot, we can use other aesthetics like colour, shape, and size or faceting. Faceting creates tables of graphics by displaying the same graph for each subset. -->
<!-- - Facet:  $B_j(s)$ -->

@@

Choice of plots are dictated by the statistical transformations and geometric objects used for the visualization. The basic plot choice for our data structure is the one that can display densities. In Section \ref{sec:measured-variable}, it is discussed that these densities could be estimated using Kernel density estimates or some descriptive statistics. The latter do not allow us to see the shape, skewness, nature of the tail or multi-modality with so much clarity as the former. However, they avoid much clutter and help us to focus on some specific properties of the data. Kernel density estimates provide more detailed information about the distribution but depend on smoothing parameters and hence can vary for the same data set. In practice, the densities are estimated reasonably with at least 30 observations. However, even with large sample size, choosing too large a value for bandwidth can cause the estimate to oversmooth the data. The summaries based on descriptive statistics remain free of any tuning parameters or Kernels, which is sometimes more desirable for exploration.

@@

When we consider the layers of grammars considered for our data structure <$A_i$, $B_j$, $v(s)$> $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$, mapping $A_i$ mapped to $x$ position and $B_j$ to facets
suggests that levels of $C_1$, that is $A_i$'s which are placed in close proximity are to be compared against each other and each level of $B_j$ represent a group. When $A_i$ mapped to facets and $B_j$ is mapped to $x$ axis, distributions are compared across levels of $B_j$ and levels if $A_i$ start acting as a group. However, the variables used in the layers are the same, the difference in mapping emphasizes different behavior of the variables.
